---
title: "Psychometric Network Analysis"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library("dplyr")
library("ggcorrplot")
library("psych")
library("bootnet")
library("psychonetrics")
library("EGAnet")
library("networktree")
```

## Setting Up R

* **R version:** R-4.4.2 for Windows (<https://cran.r-project.org/>)
* **Rtools version:** Rtools44 for Windows (<https://cran.r-project.org/bin/windows/Rtools/rtools44/rtools.html>)
* **RStudio version:** RStudio 2024.12.0+467
* **More information on EGAnet** <https://r-ega.net/>

To download the dataset:

<https://raw.githubusercontent.com/okanbulut/network/master/scr_canada.csv>

To download the R markdown file:

<https://raw.githubusercontent.com/okanbulut/network/master/network.Rmd>

To download the entire repository:

<https://github.com/okanbulut/network/archive/refs/heads/main.zip>


```{r ch1, eval=FALSE}
install.packages(c("dplyr", "ggcorrplot", "psych", "bootnet", "psychonetrics", "EGAnet", "networktree"))

library("dplyr")
library("ggcorrplot")
library("psych")
library("bootnet")
library("psychonetrics")
library("EGAnet")
library("networktree")
```

# Student Confident in Reading (SCR) in PIRLS 2021 

The dataset comes from the PIRLS website (<https://pirls2021.org/data/>).

```{r ch2}
# Students from four provinces in Canada
scr_canada <- read.csv("scr_canada.csv", header = TRUE)

# First 6 rows of the data
head(scr_canada)

# Variable names
names(scr_canada)

# Select only the SCR items
scr <- dplyr::select(scr_canada, starts_with("ASBR08"))

# Remove students with no valid responses and recode 9 as missing
scr %<>%
  dplyr::filter_all(all_vars(!is.na(.))) %>%
  dplyr::mutate_all(~na_if(., 9)) %>%
  as.data.frame()
```

Let's check the correlation matrix of the items in SCR.

```{r ch3}
# Save the correlation matrix
cormat <- psych::polychoric(x = scr)$rho

# Correlation matrix plot
ggcorrplot::ggcorrplot(corr = cormat, # correlation matrix
                       type = "lower", # print only the lower part of the matrix
                       hc.order = TRUE, # hierarchical clustering
                       show.diag = TRUE, # show the diagonal values of 1
                       lab = TRUE, # add correlation values as labels
                       lab_size = 3) # Size of the labels
```

The correlation matrix above suggests that item wording effects might be presented in the SCR scale. Let's keep the original correlation matrix and start analyzing the data using a Gaussian Graphical Model (GGM).


# Gaussian Graphical Model

```{r ch4}
network1 <- bootnet::estimateNetwork(
  data = scr, 
  corMethod = "cor_auto", # for polychoric and polyserial correlations
  default = "EBICglasso", # for estimating GGM with gLASSO and EBIC
  tuning = 0.5 # EBIC tuning parameter; set to zero for BIC model selection
)

# Print the estimated network
print(network1)

# View the estimated network
plot(network1) 
```

Let's also try the same model using psychonetrics.

```{r ch5}
# Save item names
obsvars <- colnames(scr)

network2 <- psychonetrics::ggm(scr, vars = obsvars) %>%
  psychonetrics::runmodel()

# Check out model fit
network2 %>% psychonetrics::fit()
```

We can prune this model to remove insignificant edges.

```{r ch6}
network3 <- psychonetrics::ggm(scr, vars = obsvars) %>%
  psychonetrics::runmodel() %>%
  psychonetrics::prune(adjust = "BH", alpha = 0.05)

# View the model parameters
network3 %>% psychonetrics::parameters()

# Look at the model fit
network3 %>% psychonetrics::fit()

# Compare the models
comparison <- psychonetrics::compare(
  `1. Original model`  = network2,
  `2. Sparse Model: Only Pruning` = network3)

print(comparison)
```

# Exploratory Graph Analysis

```{r ch7}
# Dimension stability analysis via EGAnet
bootEGA1 <- EGAnet::bootEGA(
  data = scr, 
  cor = "cor_auto",
  uni.method = "louvain",
  iter = 500, # Number of replica samples to generate
  # resampling" for n random subsamples of the original data
  # parametric" for n synthetic samples from multivariate normal dist.
  type = "parametric", 
  # EGA Uses standard exploratory graph analysis
  # EGA.fit Uses total entropy fit index (tefi) to determine best fit of EGA
  # hierEGA Uses hierarchical exploratory graph analysis
  EGA.type = "EGA", 
  model = "glasso", 
  algorithm = "walktrap", # or "louvain" (better for unidimensional structures)
  # use "highest_modularity", "most_common", or "lowest_tefi"
  consensus.method = "highest_modularity", 
  typicalStructure = TRUE, # typical network of partial correlations
  plot.typicalStructure = TRUE, # returns a plot of the typical network
  ncores = 8, # Number of cores to use in computing results
  seed = 2024 # set the seed for replicability
)

# View the number of communities
bootEGA1$EGA
bootEGA1$typicalGraph$typical.dim.variables

# Dimension (i.e., structural) stability results
dim_scr <- EGAnet::dimensionStability(bootEGA1)
dim_scr$dimension.stability

# Item stability results
dim_scr$item.stability
dim_scr$item.stability$plot # to see only the plot
```


# Random-Intercept EGA

Random-Intercept EGA estimates the number of dimensions after controlling for wording effects. EGA is applied to a residual correlation matrix after subtracting a random intercept factor model with equal unstandardized loadings from all the regular and unrecoded reversed items in the data.

```{r ch8}
riEGA <- EGAnet::bootEGA(
  scr, 
  uni.method = "LE",
  iter = 500,
  type = "parametric",
  cor = "cor_auto",
  model = "glasso",
  EGA.type = "riEGA", # select random-intercept EGA here
  consensus.method	= "highest_modularity", 
  consensus.iter = 100,
  algorithm="walktrap",
  seed = 2024
)

print(riEGA)
```

# Network Tree Analysis

In the final step, we will analyze whether the network structure of the SCR scale differs by gender and reading achievement (i.e., above or below the mean reading achievement score) using network tree analysis. 

```{r ch9}
scr2 <- dplyr::select(scr_canada, starts_with("ASBR08"), IDCNTRY, ITSEX, ASRREA01) %>%
  dplyr::filter_all(all_vars(!is.na(.))) %>%
  dplyr::mutate_all(~na_if(., 9)) %>%
  dplyr::mutate(gender = as.factor(ifelse(ITSEX==1,"girl", 
                                          ifelse(scr_canada$ITSEX==2, "boy", NA))),
                reading = as.factor(ifelse(ASRREA01 < 500, "below average", "above average")),
                province = as.factor(ifelse(IDCNTRY == 9130, "Newfoundland and Labrador", ifelse(IDCNTRY == 9133, "Quebec", ifelse(IDCNTRY == 9134, "Alberta", "British Columbia"))))) %>%
  as.data.frame()


nmt_scr <- networktree::networktree(nodevars=scr, 
                                    splitvars=scr2[,c("gender", "reading", "province")],
                                    transform="pcor",
                                    na.action=na.omit)

plot_scr <- plot(nmt_scr, transform="pcor", maximum=0.2, edge.width=5, vsize=10, theme="colorblind", tnex = 3, partyargs=list(ep_args = list(justmin = 15), gp = grid::gpar(cex = .5)))
```

